---
title: "Date and Time"
author: "Adithi R Upadhya"
date: "2022-08-10"
output: 
  html_document:
    toc: true
    toc_float: true
    code_download: true
---

## Load Packages

```{r warning = FALSE, message = FALSE}
library(tidyverse) 
# install.packages("lubridate")
library(lubridate)
library(nycflights13)
library(openair)
```

## Get today's date and time 

```{r}
today()

now()
```

## Create a date time object from a string 

```{r}
ymd("2022-01-31")
mdy("January 31st, 2022")
dmy("31-Jan-2022")

ymd_hms("2022-01-31 20:11:59")
```

## Create a date time object from individual components


```{r}
glimpse(flights)
new_flights <- flights %>% 
  select(year, month, day, hour, minute)

new_flights_date <- new_flights %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))
```


## From other types

```{r}
as_datetime(now())
as_date(now())
```

## Quiz 1

```{r}
d1 <- mdy("January 1, 2010")

d2 <- ymd("2015-Mar-07")

d3 <- dmy("06-Jun-2017")

d4 <- mdy(c("August 19 (2015)", "July 1 (2015)"))

d5 <- mdy("12/30/14") 
```

## Quiz 2

```{r}
ymd(c("2010-10-10", "bananas"))
```

## Manipulating date and time 

```{r}
datetime <- ymd_hms("2022/Jul/08 12:34:56")
year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
```


## Date time components

```{r}
month(datetime, label = TRUE, abbr = FALSE)
wday(datetime)
```

```{r}
hour(datetime)
minute(datetime)
second(datetime)
```

## Rounding

```{r}
datetime <- "2022-Jul-08 12:34:56"
datetime <- ymd_hms(datetime)
ceiling_date(datetime, "min")
floor_date(datetime, "hour")
round_date(datetime, "month")
```

## Setting components

```{r}
(datetime <- ymd_hms("2020-Jul-08 12:34:56"))

year(datetime) <- 2022
datetime

month(datetime) <- 08
datetime

hour(datetime) <- hour(datetime) + 2
datetime

```

## Using `update()` 

```{r}
(datetime <- ymd_hms("2020-Jul-08 12:34:56"))

update(datetime, year = 2022, month = 08, mday = 12, hour = 21)

```

## Quiz 3 

```{r}
# install.packages("openair")
library(openair)
our_new_dataset <- openair::mydata %>% 
  head(50)
```


```{r}
our_new_mutated_dataset <- our_new_dataset %>% 
  mutate(year_new_column = year(date), 
         month_new_column = month(date),
         day_new_column = mday(date))

year(our_new_mutated_dataset$date) <- 2022
```

## Durations 

```{r}
my_age <- today() - ymd("1995-03-27")
my_age
as.duration(my_age)
```

## Different durations 

```{r}
dseconds(15)
dminutes(10)
dhours(c(12, 24))
ddays(0:5)
dweeks(3)
dyears(1)
```

## Airthmetic with Durations

```{r}
2 * dyears(1)
dyears(1) + dweeks(12) + dhours(15)
today() + ddays(1)
today() - dyears(1)
```

## Periods 

```{r}
one_pm <- ymd_hms("2022-04-12 13:00:00", tz = "America/New_York")
one_pm + ddays(1)
```

## Manipulating Periods 


```{r}
seconds(15)
hours(c(12, 24))
days(7)
months(1:6)
weeks(3)
```

## Airthmetic with Periods

```{r}
10 * (months(6) + days(1))
ymd("2020-01-01") + dyears(1)
```

## Intervals 

```{r}
next_year <- today() + years(1)
(today() %--% next_year) / ddays(1)
```

## Available time zones

```{r}
Sys.timezone()
head(OlsonNames())
```

## Set time zones

```{r}
xa <- with_tz("2022-06-02 02:30:00", tzone = "Asia/Kolkata")
xa
xb <- force_tz(xa, tzone = "America/New_York")
xb
xc <- mdy_hms("09/12/2021 09:00:10", tz = "Asia/Kolkata")
xc
```

## Quiz 5.1

```{r message = FALSE, warning = FALSE}
join_data_1 <- read_csv(here::here("data", "join_data1.csv"))
join_data_2 <- read_csv(here::here("data", "join_data2.csv"))
```


```{r message = FALSE, warning = FALSE, eval = FALSE}
summary(join_data_1)
summary(join_data_2)
```

## Quiz 5.3

```{r message = FALSE, warning = FALSE}
join_data_1$date <- ymd_hms(join_data_1$date, tz = "Asia/Kolkata")
join_data_2$Date <- dmy_hms(join_data_2$Date, tz = "Asia/Kolkata")
```

## Quiz 5.4

```{r message = FALSE, warning = FALSE}
new_joined_df <- full_join(join_data_1, join_data_2, by = c("date" = "Date", 
                                                            "area" = "Area"))
```

## Quiz 5.5

```{r message = FALSE, warning = FALSE}
new_joined_df <- new_joined_df %>% 
  select(everything(), - so2)
```

## Quiz 5.6

```{r message = FALSE, warning = FALSE}
new_joined_df <- new_joined_df %>% 
  filter(area == "BLR")
```


## Quiz 5.7

```{r message = FALSE, warning = FALSE}
new_joined_df <- new_joined_df %>% 
  mutate(ratio = pm25 / pm10)
```

## Quiz 5.8

```{r message = FALSE, warning = FALSE}
 new_joined_df %>% 
  ggplot(aes(pm25, pm10)) + 
  geom_point()
```

## Quiz 5.9

```{r message = FALSE, warning = FALSE}
 new_joined_df %>% 
  ggplot(aes(date, ratio)) + 
  geom_line()
```



