---
title: "Date and Time"
author: "Adithi R. Upadhya"
date: "11 August, 2022"
format: 
  revealjs:
    theme: [ppt_theme.scss]
    incremental: true
    code-copy: true
   
execute:
  echo: true
  cache: true
---

## Recap

::: {.nonincremental}

Mutating joins, filtering joins, and set operations. 
:::

::: footer
[Learn more](https://www.tidyverse.org/packages/)
:::



## Prerequsities

```{r warning = FALSE, message = FALSE}
library(tidyverse) 
library(lubridate)
library(nycflights13)
library(openair)
```

::: footer
[Learn more](https://www.tidyverse.org/packages/)
:::

## Shortcut of the day

`Crtl + Shift + N`

</br>

. . .

::: {.fragment .highlight-red}
Opens a new document.
:::

## Star of the day `lubridate`

::: {.nonincremental}

`lubridate` is an R package that makes it easier to work with dates and times.

Date/times are represented in a unique class type.

:::

## Creating date/times
::: {.nonincremental}
There are three types of date/time data that refer to an instant in time:

-   A date. Tibbles print this as `<date>`.

-   A time within a day. Tibbles print this as `<time>`.

- A date-time is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as `<dttm>`. 
:::

## Today's data `nycflights13`

On-time data for a random sample of flights departing New York City airports in 2013.


## To get today's time

```{r}
today()

now()
```
## Create from a string

::: {.nonincremental}

Helpers `lubridate` in  automatically work out the format once you specify the order of the component. 

To use them, identify the order in which year, month, and day appear in your dates, then arrange 'y', 'm', and 'd' in the same order. 

:::

```{r}
ymd("2022-01-31")
mdy("January 31st, 2022")
dmy("31-Jan-2022")

ymd_hms("2022-01-31 20:11:59")
```

## From individual date-time components

To create a date/time from this sort of input, use `make_date()` for dates, or `make_datetime()` for date-times.

```{r}
new_flights <- flights %>% 
  select(year, month, day, hour, minute)

new_flights_date <- flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))
```


## From individual date-time components

To create a date/time from this sort of input, use `make_date()` for dates, or `make_datetime()` for date-times.

```{r, `code-line-numbers` = "6"}
new_flights <- flights %>% 
  select(year, month, day, hour, minute)

new_flights_date <- flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))
```


##  From other types

::: {.nonincremental}

Sometimes we may want to switch between a date-time and a date. 

Then we use `as_datetime()` and `as_date()`.

:::

```{r}
as_datetime(now())
as_date(now())
```


## Quiz 1  

::: {.panel-tabset}

### Question

::: {.nonincremental}

Use the appropriate lubridate function to parse each of the following dates:

::: columns
::: {.column width="100%"}

d1 <- "January 1, 2010"

d2 <- "2015-Mar-07"

d3 <- "06-Jun-2017"

d4 <- c("August 19 (2015)", "July 1 (2015)")

d5 <- "12/30/14" # Dec 30, 2014

:::
::: 
:::

### Solution

Use the appropriate lubridate function to parse each of the following dates:

```{r}
d1 <- mdy("January 1, 2010")

d2 <- ymd("2015-Mar-07")

d3 <- dmy("06-Jun-2017")

d4 <- mdy(c("August 19 (2015)", "July 1 (2015)"))

d5 <- mdy("12/30/14") 
```


:::

## Quiz 2 

::: {.panel-tabset}

### Question

What happens when you parse a string which is not a valid date, like the one shown below:

```{r, eval = FALSE, message = FALSE, warning = FALSE}
ymd(c("2010-10-10", "bananas"))
```


### Solution

What happens when you parse a string which is not a valid date, like the one shown below:

```{r}
ymd(c("2010-10-10", "bananas"))
```

:::

## Manipulating date and time

::: {.nonincremental}

In `lubridate` you can get and set individual components. 

Pull out individual parts of the date with the accessor functions.

:::

```{r}
datetime <- ymd_hms("2022-07-08 12:34:56")
year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
```
## Date-time components

::: {.nonincremental}

Using relevant arguments will show us the month abbreviations or full names. 

Time components also extracted in the same way. 

:::

```{r}
month(datetime, label = TRUE)
wday(datetime, label = TRUE, abbr = FALSE)
```

```{r}
hour(datetime)
minute(datetime)
second(datetime)
```


## Rounding

::: {.nonincremental}

An alternative approach to plotting individual components is to round the date to a nearby unit of time, with `floor_date()`, `round_date()`, and `ceiling_date()`.

Each function takes a vector of dates to adjust and then the name of the unit round down (floor), round up (ceiling), or round to. 

```{r}
datetime <- ymd_hms("2022-07-08 12:34:56")
ceiling_date(datetime, "min")
floor_date(datetime, "hour")
round_date(datetime, "month")
```


:::


## Setting components

::: {.nonincremental}

We can use each accessor function to set the components of a date/time. 

```{r}
(datetime <- ymd_hms("2020-07-08 12:34:56"))

year(datetime) <- 2022
datetime

month(datetime) <- 08
datetime

hour(datetime) <- hour(datetime) + 2
datetime

```
:::

## Or use a new function `update()`

```{r}
(datetime <- ymd_hms("2020-07-08 12:34:56"))

update(datetime, year = 2020, month = 2, mday = 2, hour = 2)

```

## Quiz 3 

::: {.panel-tabset}

### Question

::: {.nonincremental}

Extract month, year, and day from the data shown below, make three new columns of these. Hint use mutate(). 
You will have to install package called `openair` and then load it. 

```{r}
# install.packages("openair")
library(openair)
our_new_dataset <- openair::mydata %>% 
  head(50)
```

:::

### Solution

::: {.nonincremental}

Extract month and year from the data shown below, make three new columns of these. 

```{r}
our_new_mutated_dataset <- our_new_dataset %>% 
  mutate(year_new_column = year(date), 
         month_new_column = month(date))
```

:::

:::


## Quiz 4 

::: {.nonincremental}

Using the data frame created above `our_new_mutated_dataset` set the year values to `2022` instead of `1998`. 

::: columns
::: {.column width="100%"}

a)    floor_date(our_new_mutated_dataset$date, "year")
b)    ymd(our_new_mutated_dataset$date) 

::: {.fragment .highlight-red}
c)   year(our_new_mutated_dataset$date) <- 2022 
:::

d)    year(date) <- 2022 
:::

:::

:::

## Time spans

-   Durations, which represent an exact number of seconds.

-   Periods, which represent human units like weeks and months.

-   Intervals, which represent a starting and ending point.

## Durations 

::: {.nonincremental}

-   In R if you subtract two dates you get a `difftime` object. 

-   A `difftime` class object records a time span of seconds, minutes, hours, days, or weeks. 

-   This ambiguity can make `difftimes` a little painful to work with, so `lubridate` provides an alternative which always uses seconds: the duration.

```{r}
my_age <- today() - ymd("1995-03-27")
my_age
as.duration(my_age)
```
:::


## Durations 

::: {.nonincremental}

Durations come with a bunch of convenient constructors:

```{r}
dseconds(15)
dminutes(10)
dhours(c(12, 24))
ddays(0:5)
dweeks(3)
dyears(1)
```

:::


## Airthmetic with Durations

::: {.nonincremental}

```{r}
2 * dyears(1)
dyears(1) + dweeks(12) + dhours(15)
today() + ddays(1)
today() - dyears(1)
```

:::

## Periods

::: {.nonincremental}

Periods are time spans but don’t have a fixed length in seconds, instead they work with “human” times, like days and months. 

```{r}
one_pm <- ymd_hms("2022-04-12 13:00:00", tz = "America/New_York")
one_pm + ddays(1)
```


:::

## Manipulate Periods

::: {.nonincremental}

Periods come with a bunch of convenient constructors:


```{r}
seconds(15)
hours(c(12, 24))
days(7)
months(1:6)
weeks(3)
```

:::


## Airthmetic with Periods



```{r}
10 * (months(6) + days(1))
ymd("2020-01-01") + dyears(1)
```

## Intervals

::: {.nonincremental}

An interval is a duration with a starting point: that makes it precise so you can determine exactly how long it is.

```{r}
next_year <- today() + years(1)
(today() %--% next_year) / ddays(1)
```

:::


## What to use??

::: {.nonincremental}

Pick the simplest data structure that solves your problem. 

If you only care about physical time, use a duration; if you need to add human times, use a period; if you need to figure out how long a span is in human units, use an interval.

:::

## Time zones

::: {.panel-tabset}

### Available time zones

::: {.nonincremental}

-   You can find out what R thinks your current time zone is with `Sys.timezone()`.

-   In R, the time zone is an attribute of the date-time that only controls printing. Unless otherwise specified, lubridate always uses UTC. 

```{r}
Sys.timezone()
head(OlsonNames())
```


:::

### Set time zone

::: {.nonincremental}

- Keep the instant in time the same, and change how it's displayed. Use this when the instant is correct, but you want a more natural display.

```{r}
xa <- with_tz("2022-06-02 02:30:00", tzone = "Asia/Kolkata")
xa
xb <- force_tz(xa, tzone = "America/New_York")
xb
xc <- mdy_hms("09/12/2021 09:00:10", tz = "Asia/Kolkata")
xc
```

:::

:::


## Quiz 5.1

::: {.panel-tabset}

### Question


Read two files provided in the [link](https://drive.google.com/drive/folders/1azPWXaNOyhQB1js3Ne-xt_efFeWyuYnT) and name them as join_data_1 and join_data_2 accordingly. Next also look at the various statistics of these files.

Hint `read_csv()` and `summary()`.



### Solution


```{r message = FALSE, warning = FALSE}
join_data_1 <- read_csv(here::here("data", "join_data1.csv"))
join_data_2 <- read_csv(here::here("data", "join_data2.csv"))
```

</br>



```{r message = FALSE, warning = FALSE, eval = FALSE}
summary(join_data_1)
summary(join_data_2)
```


:::



## Quiz 5.2

::: {.nonincremental}

Identify the keys in both the tables.

::: {.fragment .highlight-red}

a)    date or Date

:::

b)    area or Area

::: {.fragment .highlight-red}

c)   date and area or Date and Area

:::

d)    ...1 or ...1

:::

## Quiz 5.3 

::: {.panel-tabset}

### Question


Convert the `date` column in `join_data_1` and the `Date` column in `join_data_2` to a datetime object with the timezone as "Asia/Kolkata".


Hint `ymd_hms()` for `date` column in join_data_1 and `dmy_hms()` for `Date` column in join_data_2. Mention the timezone.



### Solution

```{r message = FALSE, warning = FALSE}
join_data_1$date <- ymd_hms(join_data_1$date, tz = "Asia/Kolkata")
join_data_2$Date <- dmy_hms(join_data_2$Date, tz = "Asia/Kolkata")
```


:::


## Quiz 5.4 

::: {.panel-tabset}

### Question

Join the two tables (`join_data_1` and `join_data_2`) to get a data frame called `new_joined_df` which contains all columns and all rows from both the tables using the key identified above. 

Hint `full_join()`.

### Solution


```{r message = FALSE, warning = FALSE}
new_joined_df <- full_join(join_data_1, join_data_2, by = c("date" = "Date", 
                                                            "area" = "Area"))
```


:::


## Quiz 5.5

::: {.panel-tabset}

### Question

Remove the `so2` column.

### Solution


```{r message = FALSE, warning = FALSE}
new_joined_df <- new_joined_df %>% 
  select(everything(), - so2)
```


:::

## Quiz 5.6 

::: {.panel-tabset}

### Question

Extract areas which match with "BLR".


Hint `filter()`.

### Solution


```{r message = FALSE, warning = FALSE}
new_joined_df <- new_joined_df %>% 
  filter(area == "BLR")
```

:::


## Quiz 5.7 

::: {.panel-tabset}


### Question

Add a new column called ratio which is equal to pm25 / pm10.

Hint `mutate()`.

### Solution


```{r message = FALSE, warning = FALSE}
new_joined_df <- new_joined_df %>% 
  mutate(ratio = pm25 / pm10)
```
:::


## Quiz 5.8 

::: {.panel-tabset}

### Question

Make a scatter plot between pm25 vs pm10 using ggplot.

Hint `geom_point()`.


### Solution


```{r message = FALSE, warning = FALSE}
 new_joined_df %>% 
  ggplot(aes(pm25, pm10)) + 
  geom_point()
```
:::



## Quiz 5.9 

::: {.panel-tabset}


### Question

Make a time-series plot of ratio using ggplot.

Hint `geom_line()`.


### Solution


```{r message = FALSE, warning = FALSE}
 new_joined_df %>% 
  ggplot(aes(date, ratio)) + 
  geom_line()
```

:::


## Resources

::: {.nonincremental}

-   R for Data Science [Chapter 16](https://r4ds.had.co.nz/dates-and-times.html)
-   Slides made using [Quarto](https://meghan.rbind.io/blog/quarto-slides/#quarto-resources)
- More on [lubridate](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html#)

:::


